%!TEX root = ../main.tex

\chapter*{Acknowledgements}

I would like to express all my gratitude to my mentors Riccardo Brugnera and Alberto
Garfagnini, for welcoming me in their research group and providing me with everything I
needed to successfully carry on my research project. Thank you for the time you dedicated
to me, especially when things were going wrong. Thank you for the intellectual honesty of
your advices, thank you for the opportunities you offered me and for helping me designing
my future as a researcher. I believe not so often a Ph.D.~student has the luck to receive
this kind of support and motivation from its research group.
\newpar
A Ph.D.~research project is never a one-man effort. This thesis would not have been
successfully completed without the guidance and the collaboration of several people.
Before everyone, I would like to thank Katharina von Sturm for being the great person she
is, both as a physicist and as a human being. Thank you for being always and
unconditionally there to help, I wish everyone could have to possibility to work side to
side with a friend. I would like to thank all friends from the M\"unchen TUM group too,
for collaborating to the work presented in this thesis and for the many fruitful
discussions.  Special thanks to Matteo Agostini, Christoph Wiesinger and Elisabetta
Bossio.
\newpar
Finally, I would like to acknowledge the help I got from Prof.~Francesco Iachello and
Dr.~Jenny Kotila, who kindly provided me with up-to-date theoretical calculations for
two-neutrino double-beta decay.

\chapendgliph{}

% vim: tw=90
